Method

### 2025-11-25 GroupPolarizationStatmod/paper/method.qmd
started on specify sig thresh

---
editor: 
  markdown: 
    wrap: 80
bibliography: this.bib
---

@@

# Method

@@

Null-polarization We thus model opinions at each time, $t=\mathrm{pre}$ and
$t=\mathrm{post}$, as random variables drawn from a normal distribution with
mean $\mu_t$ and standard deviation $\sigma_t$. we drop the time

The null-polarization distributions are represented by $\nu$.

in a simple, general way---i.e., a decrease in opinion variance but a static
opinion mean---we represent . In the general group polarization hypothesis, the
group is assumed to start out mostly agreeing in principle (i.e., mostly all
agree or disagree) but to different extremes. After discussion

Showing a replication is undecidable is sufficient to undermine it. However, it
does not tell us what is the probability that a particular experimental design
produces false positives, due to clipping or any other problem.

## Simulation of Mere Consensus and Opinion Change Measurement

We model \emph{mere consensus} and opinion change measurement in group
polarization experiments to examine whether published replications can be
explained by this particular null polarization model. We simulate consensus and
measurement two ways for complementary purposes. First is this *expectation*
model, and the second is the *stochastic* model. Both models assume group
opinions are normally distributed, which is practical due to its simplicity.

We assume that participants' true latent opinions at some time, $t$, are random
variables drawn from a normal distribution with mean $\mu_t$ and standard
deviation $\sigma_t$. A normal distribution is scientifically justifiable, since
opinions are aggregations of many constituent electrical and chemical signals in
the brain—the distribution of aggregations like this tend to be *normal* (hence
the name). A normal distribution also makes sense because several of the
analysis procedures from the corpus use a $t$-test or other basic linear
regression-style test, which implicitly assume data are distributed normally.

The two models differ in how they simulate opinions, opinion change, and opinion
change measurement. We use the *expectation model* as a computationally
efficient way to simulate the observed mean opinion as the mathematical
*expectation value* of the distribution of measured opinions. In the second
model, the *stochastic model*, we simulate participant survey reporting as
drawing a latent opinion from the latent opinion distribution, then binning that
opinion in the associated opinion bin. We use the expectation model for its
efficiency in the first step of our analysis: to find a *null-polarization
model* that also fits the data for each experimental condition studied—if we
find one that means we must retract this replication since this proves that the
group polarization is undecidable with the data we have. Once we find any such
undecidable results, we use the null-polarization model to help us further
evaluate the experimental design itself.

@@

### Latent Opinion Model

Throughout this paper we model a participant $i$'s latent opinion at time $t$ as
a normally-distributed random variable,

$$ 
\omega_{i,t} \sim
\mathcal{N}(\mu_t, \sigma_t).  
$$
Group polarization in latent opinions is defined as any pair of pre- and
post-discussion opinion distributions where the mean increases in magnitude over
time—formally this is $| \mupre | < | \mupost |$. We need to take magnitudes since radicalization may be towards increasingly negative or positive opinion values.

A *null-polarization hypothesis*, or *model*, is one where the latent mean stays
constant, i.e. $\mu = \mu_\mathrm{pre} = \mu_\mathrm{post}$. *Mere agreement* is
one type of null-polarization model, 
the variance decreases over time to model agreement as described above. The mere agreement model is the cornerstone of our analysis, since it is empirically plausible and generates *spurious* observations of group polarization under mere agreement models, due to measurement artifacts induced by clipping all latent opinion values outside the measurement range—we call this *spurious polarization* for short. We simulate opinion measurements in two different ways, but both use the mere agreement model of latent opinions to test whether and by how much the measurement procedure distorts null-polarization to look like group polarization.

### The Expectation Model of Opinion Measurement

We use the \emph{expectation model} of opinion measurement to \emph{Identify} which, if any, mere agreement null-polarization models generate spurious group polarization in a particular experiment, $e$. The trick to this model, and the source of its name, is that we equate the theoretical \emph{expected value} of opinion measurements with the simulated \emph{observed average} of participant opinions.

We simulate the measurement of the mean observed opinion at time $t$ 
by first calculating integrating the normal probability density function over measurement bin boundaries $f(\omega)$, over each bin's corresponding range of opinion values. For example, consider an ordinal scale  that runs from -3 (strongly disagree) to +3 (strongly agree) (as
used by @Moscovici1969, for example). the frequency with which the "2" bin is
chosen is just the total area underneath the probability density curve between
1.5 and 2.5.

@@

#### Example: the "Americans" condition in Moscovici and Zavalloni (1969)

@@

When we plug in the bin values and copmuted $f_\mathrm{pre}$ values into the simulated average pre-discussion opinion definition (@eq-sim-ave-observed),
$$
\bar{o}_\mathrm{pre} = -3 \cdot 0.38~~+-2\cdot 0.09~~+\ldots+3\cdot0.19 = -0.6
$$
$$
\bar{o}_\mathrm{post} = -3 \cdot 0.25~~+-2\cdot 0.19~~+\ldots+3\cdot0.03 = -1.04
$$
So, we find that for the `Moscovici1969_Americans`, our null-polarization model appears polarized, with the exact same average pre- and post-discussion opinions reported by Moscovici and Zavalloni (1969), and the same magnitude of radicalization (CHECK THE PRESENTATION OF ALL THREE ABOVE)
$$
g(\nu, e) = \bar{o}_\mathrm{post} - \bar{o}_\mathrm{pre} = -1.04 - (-0.6) = - 0.4.
$$
::: {style="text-align: center;"}
FIGURE [\[fig:distros\]](#fig:distros){reference-type="ref"
reference="fig:distros"} ABOUT HERE 
:::

The replication of the `Moscovici1969_Americans` experimental condition is therefore invalidated since we proved equifinality between the group polarization hypothesis 
(discussion induces polarization) and its null (discussion does not induce polarization). \emph{If the null fits, we must omit} the result from the set of results supporting any particular hypothesis. Therefore, despite being cited over 1,000 times as a replication, these citations are in fact erroneous—even if powerful authors hold them up as such.

But how did we actually \emph{find} the null-polarization model that generated spurious group polarization? Below we demonstrate how we use a stochastic descent-style root finding that calculates spurious group polarization for a given $\nu$, then takes a step in a random direction, biased towards the direction of less error. It is therefore essential to use this efficient calculation simulation method, since tens, hundreds, or more evaluations are required for this root-finding algorithm. We show exactly how we put this to work in the following subsection explaining the algorithm to finish the \emph{Identification} step of our analysis.

<!-- OLD STUFF BELOW BEFORE QMD REWRITE -->
<!-- Moscovici and Zavalloni report -->
<!-- (Figure [\[fig:distros\]](#fig:distros){reference-type="ref" -->
<!-- reference="fig:distros"}). First, calculate the vector of probability density -->
<!-- values for each bin, $f[b]$, for the pre- and post-discussion latent opinion -->
<!-- distributions using Equation [\[eq:\]](#eq:){reference-type="ref" -->
<!-- reference="eq:"} (calculated with -->
<!-- \['makeProbVec'\](https://github.com/mt-digital/gp-statmod/blob/final-draft/GroupPolarizationStatmod/model.R#L43) -->
<!-- from the \[project code -->
<!-- repository\](https://github.com/mt-digital/gp-statmod/tree/final-draft/GroupPolarizationStatmod) -->
## Root-finding Algorithm to Find Null-Polarization Models with Artifacts

## Corpus Analysis to Determine Undecidable Replications

To analyze published replications we need to expand our notation to index
individual *experiments* with replications, and from which different journal
articles, track and analyze it,

## Evaluating Grouop Polarization Experimental Designs

>>>>

---
title: Method
editor: 
  markdown: 
    wrap: 80
author: Matt Turner · [https://mat.phd](https://mat.phd)
date: today
toc: true
bibliography: this.bib
---

@@

@@

# Latent Opinion Model

Throughout this paper we model a participant $i$'s latent opinion at time $t$ as
a normally-distributed random variable,
$$ 
\omega_{i,t} \sim
\mathcal{N}(\mu_t, \sigma_t),
$$ {#eq-random-variable-def}
with mean $\mu_t$ and standard deviation $\sigma_t$. 
Group polarization in terms of latent opinions is then defined as any pair of pre- and
post-discussion opinion distributions where the mean increases in magnitude over
time—formally, $| \mupost | > | \mupre |$. We need to take magnitudes since radicalization may be towards increasingly negative or positive opinion values.

A *null-polarization model* is a model that is consistent with the *negation*, or *null*, of the  *group polarization hypothesis*. The group polarization hypothesis is that group discussion causes the average group opinion to increase, assuming the group is aligned on a particular issue, though opinion strength varies. Therefore a null model is one where the latent mean stays
constant, i.e., $\mu = \mu_\mathrm{pre} = \mu_\mathrm{post}$.

*Mere agreement* is the type of null-polarization model we use here. Mere agreement is a null-model because the mean latent opinion is set to be constant before and after discussion. We represent *agreement*—or equivalently, *consensus*—as a decrease in the standard deviation that defines the pre- and post-discussion latent opinion distributions. The mere agreement model is the cornerstone of our analysis, since it is empirically plausible and generates *spurious* observations of group polarization under mere agreement models, due to measurement artifacts induced by clipping all latent opinion values outside the measurement range—we call this *spurious polarization* for short. We simulate opinion measurements in two different ways, but both use the mere agreement model of latent opinions to test whether and by how much the measurement procedure distorts null-polarization to look like group polarization.

## Two Ways to Model Measurement that Extend the Latent Opinion Model

We simulate measurement in two different ways—i.e., with two different models—for the two parts of our analysis. The first
is the *expectation model*: an
efficient way to *simulate measurement of the observed average* opinion of a group, using the *mathematical expectation value* of the distribution of measured opinions as a proxy. In the second
model, the *stochastic model*, we simulate individual participant responses by first drawing an opinion for each participant from the latent opinion distribution (@eq-. The opinions are binned to the appropriate ordinal value.

We use the expectation model for its
efficiency in the first step of our analysis: to find a *null-polarization
model* that also fits the data for each experimental condition studied—if we
find one that means we must retract this replication since this proves that the
group polarization is undecidable with the data we have. Once we find any such
undecidable results, we use the null-polarization model to help us further
evaluate the experimental design itself.

@@

The measurement models are introduced in detail along with the corresponding stage of the analysis that uses them. First, the *Identification* stage that uses the *expectation* measurement model; second, the *Evaluation* stage that uses the *stochastic* measurement model.

# *Identification* of Invalid Replications

In the first stage of our analysis, the *Identification* stage, we check if we can *invalidate* a replication of group polarization achieved in some experiment/experimental condition, which we label $e$. Our invalidation strategy is to find a mere agreement null-polarization model, call it $\nu$, that predicts the exact observations reported in each experimental replication in our corpus. If one is found, then we have found a model consistent with the *negation* of the group polarization hypothesis that explains the data as well as the group polarization hypothesis itself. In other words, the hypothesis is not proved or disproved given the data, but *undecidable*: the data cannot *decide* the hypothesis one way or another because of potential measurement artifacts. Because the experiment failed to produce a valid replication, the results are effectively stripped of evidentiary value and warrant exclusion from the aggregate evidence base.

But how do we find a $\nu$ that generates spurious group polarization? And how do we even simulate the group polarization experiments and measurements that make mere agreement in latent opinions look like group polarization? We simulate 
experimental measurements of group polarization by using the mathematical *expectation value* of latent opinions defined by $\nu$ as a proxy for the observed average value of ordinal-valued reported opinions (described in the first subsection below). We then use this measurement model in a numerical root-finding algorithm that identifies a $\nu$ that matches published observations, if possible. If we find such a $\nu$, this means we have a model
consistent with the null hypothesis that explains published observations. In other words, the replication is invalidated since the group polarization hypothesis is *undecidable* given the data—i.e., it cannot decide whether the hypothesis or its null is correct.

So, to review, this subsection introduces how the *Identification* stage of the analysis works. In this stage we identify for which experiments, $e$, does a mere agreement null model look like group polarization under $e$'s measurement procedure.

## The Expectation Model of Opinion Measurement

We use the *expectation model* of opinion measurement to *Identify* which, if any, mere agreement null-polarization models generate spurious group polarization in a particular experiment, $e$. The trick to this model, and the source of its name, is that we use the theoretical *expected value* of opinion measurements as a proxy for the *observed average* of participant opinions under an experimental design, $e$.

We simulate the measurement of the mean observed opinion at time $t$ 
by first calculating integrating the normal probability density function,  $f(\omega)$, over measurement bin boundaries, over each bin's corresponding range of opinion values. For example, consider an ordinal scale  that runs from -3 (strongly disagree) to +3 (strongly agree) (as
used by @Moscovici1969, for example). the frequency with which the "2" bin is
chosen is just the total area underneath the probability density curve between
1.5 and 2.5.

@@

## Example: the "Americans" condition in Moscovici and Zavalloni (1969)

@@

When we plug in the bin values and copmuted $f_\mathrm{pre}$ values into the simulated average pre-discussion opinion definition (@eq-sim-ave-observed),
$$
\bar{o}_\mathrm{pre} = -3 \cdot 0.38~~+-2\cdot 0.09~~+\ldots+3\cdot0.19 = -0.6
$$
$$
\bar{o}_\mathrm{post} = -3 \cdot 0.25~~+-2\cdot 0.19~~+\ldots+3\cdot0.03 = -1.04
$$
So, we find that for the `Moscovici1969_Americans`, our null-polarization model appears polarized, with the exact same average pre- and post-discussion opinions reported by Moscovici and Zavalloni (1969), and the same magnitude of radicalization (CHECK THE PRESENTATION OF ALL THREE ABOVE)
$$
g(\nu, e) = \bar{o}_\mathrm{post} - \bar{o}_\mathrm{pre} = -1.04 - (-0.6) = - 0.4.
$$

::: {style="text-align: center;"}
FIGURE [\[fig:distros\]](#fig:distros){reference-type="ref"
reference="fig:distros"} ABOUT HERE 
:::

Since the simulated average opinions match the observed ones—note that as a consequence, the spurious polarization magnitude also matches the observed one.
In other words, we *identified* a null model that explains the data. 
If both a hypothesis and its null explain data equally well, then the hypothesis is *undecidable*

In this example we started with

## Root-finding Algorithm to Find Null-Polarization Models with Artifacts

# Evaluation of Experimental Designs

- The evaluation stage complements our *Identify* stage by filling in more details about the experimental design.

## False positive rate

First, how frequently do the $\nu$ that we found generate simulated data
  that fool a rigorous statistical design that properly accounts for the measurement process? This serves two roles: 
  1. Confirm that the $\nu$ we found could create simulated datasets that appear polarized under statistical analysis, which is different from using the expected value of an exact latent opinion distribution, integrated with the bin values to simulate the average opinion.
  2. Evaluate the underlying experimental design independent of the failure to account for how ordinal measurements inevitably clip extreme latent opinions.
  3. We do this by simulating experiment data using $\nu$ we identified in the previous stage. We simulate 1000 experiments indexed by $i$ for each experiment $e$.

To calculate the false detection rate we begin by simulating 
pre- and post-deliberation 
opinions by drawing $N_e$ samples from normal latent opinion
distributions that generated spurious group polarization, i.e., one sample at each experiment stage for each simulated participant. The latent opinion distributions are specified with the mere agreement null model labelled $\nu_e$.

Each trial $i$ of $N_e$ simulated pre- and post-deliberation observations is then inspected to see
if spurious group polarization occurred in the simulation. Spurious group
polarization is determined by first inferring the latent parameters of each
simulated experimental trial by fitting an ordered probit model~\cite{Liddell2018}. The procedure infers the latent distribution parameters that we in fact specified with $\nu$: $\mu_\mathrm{pre}$, $\sigma_\mathrm{pre}$, $\mu_\mathrm{post}$, and $\sigma_\mathrm{post}$.

The inferred parameters are then used to calculate the effect size for that
trial, Cohen's $d$, defined as~\cite[p. 331]{Liddell2018}
$$
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
$$ {#eq-cohens}
It is implied that these values are the inferred values, not the values from $\nu$, when used in the context of calculating Cohen's $d$. Note $d=0$ always for the true $\nu$.

The family-wise error rate is the fraction of trials 
with $d_{ei} > d^*$, written
$$
  \alpha_e = \frac{1}{N_T} \sum_i^{N_T} \Theta(d_{ei}; d^*),
$$ {#eq-study-aggregate}
where
$$
  \Theta(d_{ei}; d^*) = 
    \begin{cases}
      1 & \text{ if } d_{ei} > d^* \\
      0 & \text{ otherwise.}
    \end{cases}
$$
$d^*$ is the *significance threshold*, which must be specified by the experimenter.  There are three common significance levels which we adopt here: $d^* = 0.2$ for "low", $d^*=0.5$ for "medium", and $d^* = 0.8$ for "high".

To calculate the family-wise error rate
across all experimental conditions of a study, $s$, 
\begin{equation}
  \alpha_s = \frac{1}{|s|} \sum_{e \in s} \alpha_e,
\end{equation}
\noindent
where $e \in s$ indicates sum over all experimental conditions $e$ presented in the article associated with study $s$.
The mean across all results identified as plausibly spurious is 
\begin{equation}
  \alpha_\mathrm{all} = \frac{1}{|E|} \sum_{e \in E} \alpha_e.
\end{equation}

In the final part of our *Evaluation* analysis we adapt this analysis to re-align the question of statistical rigor: we can be specific now, and instead of wondering, "what significance level is truly rigorous", we can use our techniques to make the choice of $d^*$ to conform to a specific design requirement. Specifically we will calculate what $d^*$ must be in order to limit $\alpha \leq 0.05$ as an example.

## False discovery rate

We finally may calculate the false discovery rate by plugging this $\alpha$
into Equation~\ref{eq:fdr} and assuming empirically-motivated values for 
base rate, $b$, and statistical power, $W$, as well as more pessimistic and
optimistic settings for base rate and power.

Using the notation outlined above, the false detection rate is
\begin{equation}
  \fdr = \frac{(1 - b)\alpha}{(1 - b)\alpha + bW}.
  \label{eq:fdr}
\end{equation}

## *Specify* Significance Threshold for Low False Positive Rates

In the final analysis step we calculate what significance threshold, $d^*$,
would limit the false positive rate to 0.05 or less.

There are many possible
null models we could choose.

what significance threshold would have limited inverting this procedure to find what significance value
$d^*$ is necessary to achieve a low family-wise error rate $\alpha_e = 0.05$ for
experimental condition, $e$.

To do this we first calculate $\alpha_e(d^*)$ for a
sufficiently wide range of $d^*$. We then find the lowest $d^*$ for which
$\alpha_e(d^*) \leq 0.05$.
